Scalable and Modular Verified Compilation via Automated Proofs for Declarative Transformations

### 2025-01-31 paper.tex
Tobias' comments

A fully end-to-end verified \emph{mainstream} compiler with a small
trusted code base remains somewhat of an unattainable holy grail.
End-to-end verified compilers do exist (e.g., CompCert~\cite{leroyCompCertFormallyVerified} and CakeML~\cite{kumarCakeMLVerifiedImplementation2014}), but the extra
verification burden makes compiler development challenging, relegating
such compilers to specific high-assurance niches. On the other hand,
there are tools which can fully automatically prove the correctness of a
specific compiler transformation (e.g., Alive~\cite{lopesAlive2BoundedTranslation2021}, which has gained
widespread adoption in the LLVM community), but such tools
often rely on SMT solvers to achieve their push-button automation,
making them unsuited in our quest for a minimal trusted code base. I
wish to investigate a middle road: can we verify the core transformation
algorithm separately from specific transformation instances, and then
develop proof automation to show local
correctness of a transformation, without needing input from a formal methods export?

@@

For the optimization pass, we choose to operate on a compiler IR with the same semantics as LLVM.
Although others have formalized the semantics of LLVM IR in a proof assistant before
\cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
included no proof automation. This will, to my knowledge, be the first
effort to provide fully automated proofs for correctness of LLVM
optimizations in a way that doesn't have to trust an external solver (as
Alive~\cite{lopesAlive2BoundedTranslation2021} and even AliveInLean~\cite{leeAliveInLeanVerifiedLLVM2019} do). Removing
this dependency is important as SMT solvers are large, complex pieces of
software with a history of bugs~\cite{brummayerFuzzingDeltadebuggingSMT2009, mansurDetectingCriticalBugs2020}.

@@

Finally, to model the semantics of LLVM IR in a modular way, we'd like
to follow Zakowski et
al.~(\cite{zakowskiModularCompositionalExecutable2021}) and use
interaction trees (\cite{xiaInteractionTreesRepresenting2020}).
Unfortunately, interaction trees are \emph{coinductive} in nature, to
capture possibly non-terminating programs, and Lean currently lacks
support for coinductive type. Therefore, I'd like to continue the work
that I started in my own master's thesis~\cite{keizerImplementingDefinitionalCodatatype}, which is based on earlier
work by Avigad\etal{}~\cite{avigadDataTypesQuotients2019}, and
continue development of a Lean library for coinductive types.

@@

My research will aim to answer the following questions:
\begin{enumerate}
  \item[\textsc{RQ1}]\label{research-question-1}
    How do we fully automatically verify optimizations involving memory side effects on a mid-level compiler IR, 
    while retaining a minimal trusted code base?

@@

Now, although LLVM is a bit higher level than assembly, it's still
low-level and, e.g., features unstructured control flow. The LLVM
community realized that having more structure available would enable
even more optimizations to be done purely locally (i.e., as peephole
rewrites). Thus, MLIR (Multi Level Intermediate Representation) was
born: a framework to define compiler IRs at various levels of
abstraction. MLIR mandates that IRs are in SSA form, and takes care of
syntactic overhead. We can instantiate the framework with a specific set
of operations, called a dialect. Generally, a compiler IR is made up of
a combination of such dialects. For this chapter, we shall focus on an
IR that contains bitvector arithmetic (a fragment of the \texttt{arith}
dialect) memory manipulations (\texttt{ptr} dialect) and structured
control flow (i.e., for and while loops, \texttt{scf} dialect). 
The arithmetic and pointer operations of these dialects have exactly the same semantics as in LLVM, 
but using \emph{structured} control flow puts us just above the level of abstraction of LLVM IR.

@@

More broadly, CompCert~\cite{leroyCompCertFormallyVerified} and
CakeML~\cite{kumarCakeMLVerifiedImplementation2014} are examples
of fully end-to-end verified compilers. More recent work by Mullen et
al.~\cite{mullenVerifiedPeepholeOptimizations2016} argues for
peephole rewriting as a way to lower proof burden in CompCert
development, but they still prove these rewrites manually.

Katamaran \cite{keuchelVerifiedSymbolicExecution2022} is a
symbolic simulation tool written in Coq. However, this
project focusses on being able to run the simulator outside the ITP
environment, hence eschewing meta-programming. Our main purpose is
modular proof automation, so being tied to Lean is no problem for us,
allowing us to make different implementation choices.

@@

Implementation a verified compilation pipeline as described in this proposal could 
have an impact far beyond simply reducing the chance of bugs. It could lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimization to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.

>>>>

A fully end-to-end verified \emph{mainstream} compiler with a small
trusted code base remains somewhat of an unattainable holy grail.
End-to-end verified compilers do exist (e.g., CompCert~\cite{leroyCompCertFormallyVerified} and CakeML~\cite{kumarCakeMLVerifiedImplementation2014}), but the extra
verification burden makes compiler development challenging, relegating
such compilers to specific high-assurance niches. On the other hand,
there are tools which can fully automatically prove the correctness of a
specific compiler transformation (e.g., Alive~\cite{lopesAlive2BoundedTranslation2021}, which has gained
widespread adoption in the LLVM community), but such tools
often rely on SMT solvers to achieve their push-button automation,
making them unsuited in our quest for a minimal trusted code base. I
wish to investigate a middle road: can we verify the core transformation
algorithm separately from specific transformation instances, and then
develop proof automation to show local
correctness of a transformation, without needing a formal methods expert?

@@

For the optimization pass, we choose to operate on a compiler IR with the same semantics as LLVM.
Although others have formalized the semantics of LLVM IR in a proof assistant before
\cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
did not provide automated optimization verification. This will, to my knowledge, be the first
effort to provide fully automated proofs for correctness of LLVM
optimizations in a way that doesn't have to trust an external solver (as
Alive~\cite{lopesAlive2BoundedTranslation2021} and even AliveInLean~\cite{leeAliveInLeanVerifiedLLVM2019} do). Removing
this dependency is important as SMT solvers are large, complex pieces of
software with a history of bugs~\cite{brummayerFuzzingDeltadebuggingSMT2009, mansurDetectingCriticalBugs2020}.

@@

Finally, to model the semantics of LLVM IR in a modular way, we'd like
to follow Zakowski\etal{}~\cite{zakowskiModularCompositionalExecutable2021} and use
interaction trees~\cite{xiaInteractionTreesRepresenting2020}.
Unfortunately, interaction trees are \emph{coinductive} in nature, to
capture possibly non-terminating programs, and Lean currently lacks
support for coinductive type. Therefore, I'd like to continue the work
that I started in my own master's thesis~\cite{keizerImplementingDefinitionalCodatatype}, which is based on earlier
work by Avigad\etal{}~\cite{avigadDataTypesQuotients2019}, and
continue development of a Lean library for coinductive types.

@@

My research aims to answer the following questions:
\begin{enumerate}
  \item[\textsc{RQ1}]\label{research-question-1}
    How do we fully automatically verify optimizations involving memory side effects on a mid-level compiler IR, 
    while retaining a minimal trusted code base?

@@

Now, although LLVM is a bit higher level than assembly, it's still
low-level and, e.g., features unstructured control flow. The LLVM
community realized that having more structure available would enable
even more optimizations to be done purely locally (i.e., as peephole
rewrites). Thus, MLIR (Multi Level Intermediate Representation) was
born: a framework to define compiler IRs at various levels of
abstraction. MLIR mandates that IRs are in SSA form, and takes care of
syntactic overhead. We can instantiate the framework with a specific set
of operations, called a dialect. Generally, a compiler IR is made up of
a combination of such dialects. For this chapter, we shall focus on an
IR that contains bitvector arithmetic (a fragment of the \texttt{arith}
dialect), memory manipulations (\texttt{ptr} dialect) and structured
control flow (i.e., for and while loops, \texttt{scf} dialect). 
The arithmetic and pointer operations of these dialects have exactly the same semantics as in LLVM, 
but using \emph{structured} control flow puts us just above the level of abstraction of LLVM IR.

@@

More broadly, CompCert~\cite{leroyCompCertFormallyVerified} and
CakeML~\cite{kumarCakeMLVerifiedImplementation2014} are examples
of fully end-to-end verified compilers. More recent work by Mullen et
al.~\cite{mullenVerifiedPeepholeOptimizations2016} argues for
peephole rewriting as a way to lower proof burden in CompCert
development, but they still require manual effort to prove these rewrites.

Katamaran~\cite{keuchelVerifiedSymbolicExecution2022} is a
symbolic simulation tool written in Coq. However, this
project focusses on being able to run the simulator outside the ITP
environment, hence eschewing meta-programming. Our main purpose is
modular proof automation, so being tied to Lean is no problem for us,
allowing us to make different implementation choices.

@@

Implementing a verified compilation pipeline as described in this proposal could 
have an impact far beyond simply reducing the chance of bugs. It could lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimization to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.
